---
title: "Project 1 Final Report - Analysis Work"
author: "Madelynn Schina"
date: "March 2026"
output: pdf_document
---


```{r, include=F} 
knitr::opts_chunk$set(echo = TRUE)

library(knitr)
library(here)
library(dplyr)
library(gtsummary)
library(cmdstanr)
library(bayesplot)
library(posterior)
library(bayestestR)
library(mcmcse)
library(loo)
library(tibble)
library(kableExtra)
library(broom)
library(gt)

```


```{r}
# Loading dataset
hiv_data <- read.csv(here("Project 1/Data", "hiv_data_cleaned.csv"))

```


```{r}
# Data Management
# Ensuring necessary variables are treated as factors
hiv_data <- hiv_data %>%
  mutate(education_B = factor(education_B,
                              levels = c("Not College Graduate", "College Graduate")),
         race_B = factor(race_B,
                         levels = c("Other", "White, Non-Hispanic")),
         smoking_B = factor(smoking_B,
                            levels = c("Not Current Smoker", "Current Smoker")),
         income_B = factor(income_B,
                           levels = 1:7,
                           labels = c("Less than $10,000",
                                      "$10,000–19,999",
                                      "$20,000–29,999",
                                      "$30,000–39,999",
                                      "$40,000–49,999",
                                      "$50,000–59,999",
                                      "$60,000 or more")),
         adherence_Y2 = factor(adherence_Y2,
                               levels = c("> 95%", "< 95%")))

```




```{r, echo = F}
# Creating table 1
descriptive_table <- hiv_data %>%
  select(hard_drugs_B, VLOAD_B, LEU3N_B, AGG_PHYS_B, AGG_MENT_B,
         age_B, bmi_B, race_B, education_B, smoking_B, income_B) %>%
  tbl_summary(by = hard_drugs_B,
              missing = "ifany",
              statistic = list(all_continuous() ~ "{mean} ({sd})",
                               all_categorical() ~ "{n} ({p}%)"),
              label = list(VLOAD_B ~ "Viral Load (copies/mL)",
                           LEU3N_B ~ "CD4+ T Cell Count (# of cells)",
                           AGG_PHYS_B ~ "Physical QOL Score",
                           AGG_MENT_B ~ "Mental Health QOL Score",
                           age_B ~ "Age (years)",
                           bmi_B ~ "BMI (kg/m^2)",
                           race_B ~ "Race",
                           education_B ~ "Highest Education Reached",
                           smoking_B ~ "Smoking Status",
                           income_B ~ "Income")) %>%
  add_overall() %>%
  modify_header(stat_1 ~ "**No Hard Drug Use** <br> N = {n}",
                stat_2 ~ "**Hard Drug Use** <br> N = {n}") %>%
  modify_table_body(~ .x %>%
                      dplyr::mutate(label = ifelse(label == "Unknown", "Missing", label))) %>%
  modify_footnote(all_stat_cols() ~ "Continuous variables: Mean (SD); Categorical variables: n (%)")


# Save table as image for report
gtsave(as_gt(descriptive_table), "descriptivetable_finalreport.png")

```



## Question 1
How does treatment response 2 years after initiating HAART differ between subjects who report using hard drugs, such as heroin and cocaine, at baseline and other subjects, who did not report hard drug use at baseline?


```{r}
#### Frequentist Approach ####
# Model for Viral Load
vload_model <- lm(log_VLOAD_Y2 ~ log_VLOAD_B + age_B + bmi_B + race_B + education_B + hard_drugs_B + smoking_B + income_B, 
                  data = hiv_data)

# Model for T-Cell Count
tcell_model <- lm(LEU3N_Y2 ~ LEU3N_B + age_B + bmi_B + race_B + education_B + hard_drugs_B + smoking_B + income_B, 
                  data = hiv_data)

# Model for Mental QOL Score
ment_qol_model <- lm(AGG_MENT_Y2 ~ AGG_MENT_B + age_B + bmi_B + race_B + education_B + hard_drugs_B + smoking_B + income_B, 
                     data = hiv_data)

# Model for Physical QOL Score
phys_qol_model <- lm(AGG_PHYS_Y2 ~ AGG_PHYS_B + age_B + bmi_B + race_B + education_B + hard_drugs_B + smoking_B + income_B, 
                     data = hiv_data)

```


```{r, echo = F}
# Summary Tables for Frequentist Approach
# Results of Viral Load model
tbl_vload <- tbl_regression(vload_model,
                            conf.level = 0.95,
                            intercept = TRUE,
                            estimate_fun = ~style_number(.x, digits = 3),
                            label = list("hard_drugs_B" ~ "Hard Drugs Use Prior to Treatment",
                                          "(Intercept)" ~ "Intercept",
                                          "log_VLOAD_B" ~ "Log10 Viral Load (Baseline)",
                                          "age_B" ~ "Age",
                                          "bmi_B" ~ "BMI",
                                          "race_B" ~ "Race",
                                          "education_B" ~ "Education",
                                          "smoking_B" ~ "Smoking Status",
                                          "income_B" ~ "Income"))

# Results of T-Cell Count model
tbl_tcell <- tbl_regression(tcell_model,
                            conf.level = 0.95,
                            intercept = TRUE,
                            estimate_fun = ~style_number(.x, digits = 3),
                            label = list("hard_drugs_B" ~ "Hard Drugs Use Prior to Treatment",
                                          "(Intercept)" ~ "Intercept",
                                          "LEU3N_B" ~ "T-Cell Count (Baseline)",
                                          "age_B" ~ "Age",
                                          "bmi_B" ~ "BMI",
                                          "race_B" ~ "Race",
                                          "education_B" ~ "Education",
                                          "smoking_B" ~ "Smoking Status",
                                          "income_B" ~ "Income"))

# Results of Mental QOL Score model
tbl_ment_qol <- tbl_regression(ment_qol_model,
                               conf.level = 0.95,
                               intercept = TRUE,
                               estimate_fun = ~style_number(.x, digits = 3),
                               label = list("hard_drugs_B" ~ "Hard Drugs Use Prior to Treatment",
                                          "(Intercept)" ~ "Intercept",
                                          "AGG_MENT_B" ~ "Mental QOL Score (Baseline)",
                                          "age_B" ~ "Age",
                                          "bmi_B" ~ "BMI",
                                          "race_B" ~ "Race",
                                          "education_B" ~ "Education",
                                          "smoking_B" ~ "Smoking Status",
                                          "income_B" ~ "Income"))

# Results of Mental QOL Score model
tbl_phys_qol <- tbl_regression(phys_qol_model,
                               conf.level = 0.95,
                               intercept = TRUE,
                               estimate_fun = ~style_number(.x, digits = 3),
                               label = list("hard_drugs_B" ~ "Hard Drugs Use Prior to Treatment",
                                          "(Intercept)" ~ "Intercept",
                                          "AGG_PHYS_B" ~ "Physical QOL Score (Baseline)",
                                          "age_B" ~ "Age",
                                          "bmi_B" ~ "BMI",
                                          "race_B" ~ "Race",
                                          "education_B" ~ "Education",
                                          "smoking_B" ~ "Smoking Status",
                                          "income_B" ~ "Income"))


# Combining model results into nice table
tbl_combined <- tbl_merge(tbls = list(tbl_vload, tbl_tcell, tbl_ment_qol, tbl_phys_qol),
                          tab_spanner = c("**Viral Load Model**", "**T-Cell Model**", "**Mental QOL Model**", "**Physical QOL Model**"))


# Table for Appendix with covariates
tbl_combined

```


```{r, echo = F}
# Residual Diagnostic Plots to check transformations
par(mfrow = c(2, 2))

# Viral Load
plot(vload_model, main = "Viral Load")

# T-Cell Count
plot(tcell_model, main = "T-Cell Count")

# Mental QoL
plot(ment_qol_model, main = "Mental QoL")

# Physical QoL
plot(phys_qol_model, main = "Physical QoL")

par(mfrow = c(1, 1))
```




```{r, echo = F}
#### Bayesian Approach ####
## Step 1 - Define the model ##
# Half Normal prior on sigma and normal priors on the regression coefficients

stan_file <- write_stan_file("data {
  int<lower=0> N;                  // number of observations
  int<lower=0> P;                  // number of predictors including intercept
  matrix[N, P] X;                  // design matrix (first column = intercept)
  vector[N] y;                     // outcome

  vector[P] prior_mean;            // prior means for each beta
  vector<lower=0>[P] prior_sd;     // prior SDs for each beta

  real<lower=0> sigma_prior_sd;    // SD for half-normal prior on sigma
}

parameters {vector[P] beta;                  // regression coefficients
            real<lower=0> sigma;             // residual SD
            }

model {
  // Vectorized priors for regression coefficients
  beta ~ normal(prior_mean, prior_sd);

  // Half-normal prior for sigma
  sigma ~ normal(0, sigma_prior_sd);

  // Likelihood
  y ~ normal(X * beta, sigma);
}

generated quantities {
  // log likelihood for each observation for calculating model fit stats
  vector[N] log_lik;
  
  for (n in 1:N) {
    log_lik[n] = normal_lpdf(y[n] | X[n] * beta, sigma);}
  }", dir="STAN", basename='linear_regression_half_normal'
)


## Step 2 - Compile the Stan program ##
mod <- cmdstan_model(stan_file)


## Step 3 - Prepare your data to pass to Stan ##
# Variables of linear regression
outcomes <- c("log_VLOAD_Y2", "LEU3N_Y2", "AGG_MENT_Y2", "AGG_PHYS_Y2")
baseline_vars <- c("log_VLOAD_B", "LEU3N_B", "AGG_MENT_B", "AGG_PHYS_B")
other_covariates <- c("age_B", "bmi_B", "race_B", "education_B", "hard_drugs_B", "smoking_B", "income_B")

# Initializing list to store data for stan models
stan_data_list <- list()

for (i in seq_along(outcomes)) {
  outcome <- outcomes[i]
  baseline <- baseline_vars[i]
  
  # Select relevant columns and drop NAs
  model_vars <- c(outcome, baseline, other_covariates)
  model_data <- hiv_data[, model_vars] %>% na.omit()
  
  # Create design matrix and get y with correct dimensions without na's
  X <- model.matrix(~ ., data = model_data[, c(baseline, other_covariates)])
  y <- model_data[[outcome]]
  
  
  # Additional Variables Needed
  N <- nrow(X)      # N is the number of observations in your data set 
  P <- ncol(X)      # P is the number of columns in the design matrix
  
  # Hyperparameters for prior distributions.
  m <- c(1, rep(0, (P-1)))      # mvnorm mean
  s <- rep(100,P)               # variance 100^2
  sigma_sd <- 100
  
  # Save data for each model to list
  stan_data_list[[outcome]] <- list(N = N,
                                    P = P,
                                    X = X,
                                    y = y,
                                    prior_mean = m,
                                    prior_sd = s,
                                    sigma_prior_sd = sigma_sd)
}

```


```{r}
## Step 4 - Fitting Models ##
# Model for Viral Load
vload_fit <- mod$sample(data = stan_data_list$log_VLOAD_Y2,
                        chains = 4,
                        iter_warmup = 5000,
                        iter_sampling = 25000,
                        seed = 6624)

# Model for T-Cell Count
tcell_fit <- mod$sample(data = stan_data_list$LEU3N_Y2,
                        chains = 4,
                        iter_warmup = 5000,
                        iter_sampling = 25000,
                        seed = 6624)

# Model for Mental QOL Score
ment_qol_fit <- mod$sample(data = stan_data_list$AGG_MENT_Y2,
                           chains = 4,
                           iter_warmup = 5000,
                           iter_sampling = 25000,
                           seed = 6624)

# Model for Physical QOL Score
phys_qol_fit <- mod$sample(data = stan_data_list$AGG_PHYS_Y2,
                           chains = 4,
                           iter_warmup = 5000,
                           iter_sampling = 25000,
                           seed = 6624)

```


```{r}
## Step 5 - Summarize the posterior ##
# Defining function from code provided
summarize_posterior <- function(fit, variables = NULL, ci = 0.95) {
  # Quick summary
  if (!is.null(variables)) {
    print(as.data.frame(fit$summary(variables = variables)))
  }
  
  # Extract draws
  draws <- fit$draws(format = "array")
  draws_mat <- matrix(as.numeric(draws), nrow = dim(draws)[1] * dim(draws)[2], ncol = dim(draws)[3])
  colnames(draws_mat) <- dimnames(draws)[[3]]
  
  # Only keep parameter columns (exclude lp__ and log_lik)
  params <- colnames(draws_mat)
  params <- params[!grepl("lp__|log_lik", params)]
  
  # Summarize each parameter
  summary_table <- lapply(params, function(p) {
    vals <- as.numeric(draws_mat[, p])
    
    # Monte Carlo standard error
    mcse_val <- mcmcse::mcse(vals)$se
    
    # Effective sample size
    ess_val <- ess_bulk(vals)
    
    # 95% HPDI
    hpd <- hdi(vals, ci = ci)
    
    tibble(Parameter = p,
           Estimate  = mean(vals),
           Std_Dev   = sd(vals),
           MCSE      = mcse_val,
           HPDI_2.5  = hpd$CI_low,
           HPDI_97.5 = hpd$CI_high,
           ESS       = ess_val)
  }) %>% bind_rows()
  
  # Optional: check MCSE rule of thumb (< 6% of posterior SD)
  summary_table <- summary_table %>%
    mutate(MCSE_pct = 100 * MCSE / Std_Dev,
           MCSE_ok  = MCSE_pct < 6)
  
  return(summary_table)
}


# Applying function to all four models
posterior_summary_vload <- summarize_posterior(vload_fit, 
                                               variables = c("beta[7]"))
posterior_summary_tcell <- summarize_posterior(tcell_fit, 
                                               variables = c("beta[7]"))
posterior_summary_ment_qol <- summarize_posterior(ment_qol_fit, 
                                              variables = c("beta[7]"))
posterior_summary_phys_qol <- summarize_posterior(phys_qol_fit, 
                                              variables = c("beta[7]"))

print(posterior_summary_vload)
print(posterior_summary_tcell)
print(posterior_summary_ment_qol)
print(posterior_summary_phys_qol)

```


```{r}
## Step 6 - Model fit statistics ##
# Extract the log likelihood info
loglik_mat_vload <- as_draws_matrix(vload_fit$draws("log_lik"))
loglik_mat_tcell <- as_draws_matrix(tcell_fit$draws("log_lik"))
loglik_mat_ment <- as_draws_matrix(ment_qol_fit$draws("log_lik"))
loglik_mat_phys <- as_draws_matrix(phys_qol_fit$draws("log_lik"))

# Get LOO-CV and WAIC
loo_res_vload <- loo(loglik_mat_vload)
loo_res_tcell <- loo(loglik_mat_tcell)
loo_res_ment_qol <- loo(loglik_mat_ment)
loo_res_phys_qol <- loo(loglik_mat_phys)

waic_res_vload <- waic(loglik_mat_vload)
waic_res_tcell <- waic(loglik_mat_tcell)
waic_res_ment_qol <- waic(loglik_mat_ment)
waic_res_phys_qol <- waic(loglik_mat_phys)

print(loo_res_vload)
print(loo_res_tcell)
print(loo_res_ment_qol)
print(loo_res_phys_qol)
print(waic_res_vload)
print(waic_res_tcell)
print(waic_res_ment_qol)
print(waic_res_phys_qol)

```




```{r, echo = F}
## Step 7 - MCMC Diagnostics ##
# Defining function from code provided
generate_diagnostics <- function(fit, variables = NULL) {
 draws_array <- as_draws_array(fit$draws())
 
 # Trace plots
 mcmc_trace(draws_array, pars = variables)
 
 # R-hat and ESS diagnostics
 summary_table <- fit$summary(variables = variables)
 
 # Posterior Density Plots
 mcmc_dens_overlay(draws_array, pars = variables)
 
 # Auto-correlation 
 mcmc_acf(draws_array, pars = variables)

 # Diagnostics from cmdstan
 cmdstan_diag <- fit$cmdstan_diagnose()
 
 return(list(draws_array = draws_array,
             draws_df = draws_df,
             summary = summary_table,
             cmdstan_diag = cmdstan_diag))
}


# Applying function to all four models
diag_results_vload <- generate_diagnostics(vload_fit, variables = c("beta[7]"))
diag_results_tcell <- generate_diagnostics(tcell_fit, variables = c("beta[7]"))
diag_results_ment <- generate_diagnostics(ment_qol_fit, variables = c("beta[7]"))
diag_results_phys <- generate_diagnostics(phys_qol_fit, variables = c("beta[7]"))


# Saving output for each model
for (model_name in c("vload", "tcell", "ment_qol", "phys_qol")) {
  fit <- get(paste0(model_name, "_fit"))
  summary <- get(paste0("posterior_summary_", model_name))
  loo_res <- get(paste0("loo_res_", model_name))
  waic_res <- get(paste0("waic_res_", model_name))
  
  sink(paste0("Project 1/Code/Output/", model_name, "_MCMCOutput.txt"))
  print(summary)
  print(loo_res)
  print(waic_res)
  print(fit$cmdstan_diagnose())
  sink()
  
  pdf(paste0("Project 1/Code/Output/", model_name, "_DiagnosticPlots.pdf"))
  draws <- as_draws_array(fit$draws())
  print(mcmc_trace(draws, pars = "beta[7]"))
  print(mcmc_dens_overlay(draws, pars = "beta[7]"))
  print(mcmc_acf(draws, pars = "beta[7]"))
  dev.off()
}

```



```{r, echo = F}
# Creating table to illustrate comparison of the two appraoches
# Extract frequentist results
freq_results <- bind_rows(tidy(vload_model, conf.int = TRUE) %>% filter(term == "hard_drugs_B") %>% mutate(Outcome = "Viral Load"),
                          tidy(tcell_model, conf.int = TRUE) %>% filter(term == "hard_drugs_B") %>% mutate(Outcome = "T-Cell Count"),
                          tidy(ment_qol_model, conf.int = TRUE) %>% filter(term == "hard_drugs_B") %>% mutate(Outcome = "Mental QoL"),
                          tidy(phys_qol_model, conf.int = TRUE) %>% filter(term == "hard_drugs_B") %>% mutate(Outcome = "Physical QoL")) %>%
  mutate(Beta_freq = round(estimate, 2),
         CI_freq = paste0("(", round(conf.low, 2), ", ", round(conf.high, 2), ")"),
         pval = ifelse(p.value < 0.001, "<0.001", as.character(round(p.value, 3)))) %>%
  select(Outcome, Beta_freq, CI_freq, pval)

# Extract Bayesian results
# Calculate posterior probabilities P(beta < 0)
calc_post_prob <- function(fit, param = "beta[7]") {
  draws <- as_draws_matrix(fit$draws())
  beta_draws <- draws[, param]
  round(mean(beta_draws < 0), 3)
}

pp_vload <- calc_post_prob(vload_fit)
pp_tcell <- calc_post_prob(tcell_fit)
pp_ment <- calc_post_prob(ment_qol_fit)
pp_phys <- calc_post_prob(phys_qol_fit)

bayes_results <- bind_rows(
  posterior_summary_vload %>% filter(Parameter == "beta[7]") %>% mutate(Outcome = "Viral Load"),
  posterior_summary_tcell %>% filter(Parameter == "beta[7]") %>% mutate(Outcome = "T-Cell Count"),
  posterior_summary_ment_qol %>% filter(Parameter == "beta[7]") %>% mutate(Outcome = "Mental QoL"),
  posterior_summary_phys_qol %>% filter(Parameter == "beta[7]") %>% mutate(Outcome = "Physical QoL")
) %>%
  mutate(
    Beta_bayes = round(Estimate, 2),
    CI_bayes = paste0("(", round(HPDI_2.5, 2), ", ", round(HPDI_97.5, 2), ")"),
    Post_Prob = c(pp_vload, pp_tcell, pp_ment, pp_phys)) %>%
  select(Outcome, Beta_bayes, CI_bayes, Post_Prob)

# Combine side by side
combined <- left_join(freq_results, bayes_results, by = "Outcome")

# Table
combined %>%
  kbl(col.names = c("Outcome", "β", "95% CI", "p-value", "Posterior Mean", "95% HPDI", "P(β < 0)"),
      align = c("l", "r", "r", "r", "r", "r", "r")) %>%
  kable_classic(full_width = FALSE) %>%
  add_header_above(c(" " = 1, "Frequentist" = 3, "Bayesian" = 3)) %>%
  footnote(general = "Viral load was log10 transformed prior to analysis.",
           general_title = "",
           footnote_as_chunk = TRUE)

```




## Question 2
Could hard drug use explain the difference in adherence?

Will be fitting the same models again with adherence at year 2 as a predictor now.


```{r, echo = F}
#### Frequentist Approach ####
# Model for Viral Load with adherence
vload_model_adh <- lm(log_VLOAD_Y2 ~ log_VLOAD_B + age_B + bmi_B + race_B + education_B + hard_drugs_B + smoking_B + income_B + adherence_Y2, 
                  data = hiv_data)

# Model for T-Cell Count
tcell_model_adh <- lm(LEU3N_Y2 ~ LEU3N_B + age_B + bmi_B + race_B + education_B + hard_drugs_B + smoking_B + income_B + adherence_Y2, 
                  data = hiv_data)

# Model for Mental QOL Score
ment_qol_model_adh <- lm(AGG_MENT_Y2 ~ AGG_MENT_B + age_B + bmi_B + race_B + education_B + hard_drugs_B + smoking_B + income_B + adherence_Y2, 
                     data = hiv_data)

# Model for Physical QOL Score
phys_qol_model_adh <- lm(AGG_PHYS_Y2 ~ AGG_PHYS_B + age_B + bmi_B + race_B + education_B + hard_drugs_B + smoking_B + income_B + adherence_Y2, 
                     data = hiv_data)

```




```{r, echo = F}
## Step 3 - Prepare your data to pass to Stan ##
# Variables of linear regression
outcomes <- c("log_VLOAD_Y2", "LEU3N_Y2", "AGG_MENT_Y2", "AGG_PHYS_Y2")
baseline_vars <- c("log_VLOAD_B", "LEU3N_B", "AGG_MENT_B", "AGG_PHYS_B")
other_covariates <- c("age_B", "bmi_B", "race_B", "education_B", "hard_drugs_B", "smoking_B", "income_B", "adherence_Y2")

# Initializing list to store data for stan models
stan_data_list <- list()

for (i in seq_along(outcomes)) {
  outcome <- outcomes[i]
  baseline <- baseline_vars[i]
  
  # Select relevant columns and drop NAs
  model_vars <- c(outcome, baseline, other_covariates)
  model_data <- hiv_data[, model_vars] %>% na.omit()
  
  # Create design matrix and get y with correct dimensions without na's
  X <- model.matrix(~ ., data = model_data[, c(baseline, other_covariates)])
  y <- model_data[[outcome]]
  
  
  # Additional Variables Needed
  N <- nrow(X)      # N is the number of observations in your data set 
  P <- ncol(X)      # P is the number of columns in the design matrix
  
  # Hyperparameters for prior distributions.
  m <- c(1, rep(0, (P-1)))      # mvnorm mean
  s <- rep(100,P)               # variance 100^2
  sigma_sd <- 100
  
  # Save data for each model to list
  stan_data_list[[outcome]] <- list(N = N,
                                    P = P,
                                    X = X,
                                    y = y,
                                    prior_mean = m,
                                    prior_sd = s,
                                    sigma_prior_sd = sigma_sd)
}

```


```{r}
## Step 4 - Fitting Models ##
# Model for Viral Load
vload_fit_adh <- mod$sample(data = stan_data_list$log_VLOAD_Y2,
                            chains = 4,
                            iter_warmup = 5000,
                            iter_sampling = 25000,
                            seed = 6624)

# Model for T-Cell Count
tcell_fit_adh <- mod$sample(data = stan_data_list$LEU3N_Y2,
                            chains = 4,
                            iter_warmup = 5000,
                            iter_sampling = 25000,
                            seed = 6624)

# Model for Mental QOL Score
ment_qol_fit_adh <- mod$sample(data = stan_data_list$AGG_MENT_Y2,
                               chains = 4,
                               iter_warmup = 5000,
                               iter_sampling = 25000,
                               seed = 6624)

# Model for Physical QOL Score
phys_qol_fit_adh <- mod$sample(data = stan_data_list$AGG_PHYS_Y2,
                               chains = 4,
                               iter_warmup = 5000,
                               iter_sampling = 25000,
                               seed = 6624)

```


```{r}
## Step 5 - Summarize the posterior ##
# Defining function from code provided
summarize_posterior <- function(fit, variables = NULL, ci = 0.95) {
  # Quick summary
  if (!is.null(variables)) {
    print(as.data.frame(fit$summary(variables = variables)))
  }
  
  # Extract draws
  draws <- fit$draws(format = "array")
  draws_mat <- matrix(as.numeric(draws), nrow = dim(draws)[1] * dim(draws)[2], ncol = dim(draws)[3])
  colnames(draws_mat) <- dimnames(draws)[[3]]
  
  # Only keep parameter columns (exclude lp__ and log_lik)
  params <- colnames(draws_mat)
  params <- params[!grepl("lp__|log_lik", params)]
  
  # Summarize each parameter
  summary_table <- lapply(params, function(p) {
    vals <- as.numeric(draws_mat[, p])
    
    # Monte Carlo standard error
    mcse_val <- mcmcse::mcse(vals)$se
    
    # Effective sample size
    ess_val <- ess_bulk(vals)
    
    # 95% HPDI
    hpd <- hdi(vals, ci = ci)
    
    tibble(Parameter = p,
           Estimate  = mean(vals),
           Std_Dev   = sd(vals),
           MCSE      = mcse_val,
           HPDI_2.5  = hpd$CI_low,
           HPDI_97.5 = hpd$CI_high,
           ESS       = ess_val)
  }) %>% bind_rows()
  
  # Optional: check MCSE rule of thumb (< 6% of posterior SD)
  summary_table <- summary_table %>%
    mutate(MCSE_pct = 100 * MCSE / Std_Dev,
           MCSE_ok  = MCSE_pct < 6)
  
  return(summary_table)
}


# Applying function to all four models
posterior_summary_vload_adh <- summarize_posterior(vload_fit_adh, 
                                                   variables = c("beta[7]"))
posterior_summary_tcell_adh <- summarize_posterior(tcell_fit_adh, 
                                                   variables = c("beta[7]"))
posterior_summary_ment_qol_adh <- summarize_posterior(ment_qol_fit_adh, 
                                                      variables = c("beta[7]"))
posterior_summary_phys_qol_adh <- summarize_posterior(phys_qol_fit_adh, 
                                                      variables = c("beta[7]"))

print(posterior_summary_vload_adh)
print(posterior_summary_tcell_adh)
print(posterior_summary_ment_qol_adh)
print(posterior_summary_phys_qol_adh)

```


```{r}
## Step 6 - Model fit statistics ##
# Extract the log likelihood info
loglik_mat_vload_adh <- as_draws_matrix(vload_fit$draws("log_lik"))
loglik_mat_tcell_adh <- as_draws_matrix(tcell_fit$draws("log_lik"))
loglik_mat_ment_adh <- as_draws_matrix(ment_qol_fit$draws("log_lik"))
loglik_mat_phys_adh <- as_draws_matrix(phys_qol_fit$draws("log_lik"))

# Get LOO-CV and WAIC
loo_res_vload_adh <- loo(loglik_mat_vload)
loo_res_tcell_adh <- loo(loglik_mat_tcell)
loo_res_ment_qol_adh <- loo(loglik_mat_ment)
loo_res_phys_qol_adh <- loo(loglik_mat_phys)

waic_res_vload_adh <- waic(loglik_mat_vload)
waic_res_tcell_adh <- waic(loglik_mat_tcell)
waic_res_ment_qol_adh <- waic(loglik_mat_ment)
waic_res_phys_qol_adh <- waic(loglik_mat_phys)

print(loo_res_vload_adh)
print(loo_res_tcell_adh)
print(loo_res_ment_qol_adh)
print(loo_res_phys_qol_adh)
print(waic_res_vload_adh)
print(waic_res_tcell_adh)
print(waic_res_ment_qol_adh)
print(waic_res_phys_qol_adh)

```




```{r, echo = F}
## Step 7 - MCMC Diagnostics ##
# Defining function from code provided
generate_diagnostics <- function(fit, variables = NULL) {
 draws_array <- as_draws_array(fit$draws())
 
 # Trace plots
 mcmc_trace(draws_array, pars = variables)
 
 # R-hat and ESS diagnostics
 summary_table <- fit$summary(variables = variables)
 
 # Posterior Density Plots
 mcmc_dens_overlay(draws_array, pars = variables)
 
 # Auto-correlation 
 mcmc_acf(draws_array, pars = variables)

 # Diagnostics from cmdstan
 cmdstan_diag <- fit$cmdstan_diagnose()
 
 return(list(draws_array = draws_array,
             draws_df = draws_df,
             summary = summary_table,
             cmdstan_diag = cmdstan_diag))
}


# Applying function to all four models
diag_results_vload_adh <- generate_diagnostics(vload_fit_adh, variables = c("beta[7]"))
diag_results_tcell_adh <- generate_diagnostics(tcell_fit_adh, variables = c("beta[7]"))
diag_results_ment_adh <- generate_diagnostics(ment_qol_fit_adh, variables = c("beta[7]"))
diag_results_phys_adh <- generate_diagnostics(phys_qol_fit_adh, variables = c("beta[7]"))


# Saving output for each model
for (model_name in c("vload", "tcell", "ment_qol", "phys_qol")) {
  fit <- get(paste0(model_name, "_fit_adh"))
  summary <- get(paste0("posterior_summary_", model_name, "_adh"))
  loo_res <- get(paste0("loo_res_", model_name, "_adh"))
  waic_res <- get(paste0("waic_res_", model_name, "_adh"))
  
  sink(paste0("Project 1/Code/Output/", model_name, "_adh_MCMCOutput.txt"))
  print(summary)
  print(loo_res)
  print(waic_res)
  print(fit$cmdstan_diagnose())
  sink()
  
  pdf(paste0("Project 1/Code/Output/", model_name, "_adh_DiagnosticPlots.pdf"))
  draws <- as_draws_array(fit$draws())
  print(mcmc_trace(draws, pars = "beta[7]"))
  print(mcmc_dens_overlay(draws, pars = "beta[7]"))
  print(mcmc_acf(draws, pars = "beta[7]"))
  dev.off()
}

```


```{r, echo = F}
# Creating table for results from all models with and without adherence
# Frequentist Results
combined_adh <- bind_rows(tidy(vload_model, conf.int = TRUE) %>% filter(term == "hard_drugs_B") %>% mutate(Outcome = "Viral Load", Group = "Without Adherence"),
                          tidy(tcell_model, conf.int = TRUE) %>% filter(term == "hard_drugs_B") %>% mutate(Outcome = "T-Cell Count", Group = "Without Adherence"),
                          tidy(ment_qol_model, conf.int = TRUE) %>% filter(term == "hard_drugs_B") %>% mutate(Outcome = "Mental QoL", Group = "Without Adherence"),
                          tidy(phys_qol_model, conf.int = TRUE) %>% filter(term == "hard_drugs_B") %>% mutate(Outcome = "Physical QoL", Group = "Without Adherence"),
                          tidy(vload_model_adh, conf.int = TRUE) %>% filter(term == "hard_drugs_B") %>% mutate(Outcome = "Viral Load", Group = "With Adherence"),
                          tidy(tcell_model_adh, conf.int = TRUE) %>% filter(term == "hard_drugs_B") %>% mutate(Outcome = "T-Cell Count", Group = "With Adherence"),
                          tidy(ment_qol_model_adh, conf.int = TRUE) %>% filter(term == "hard_drugs_B") %>% mutate(Outcome = "Mental QoL", Group = "With Adherence"),
                          tidy(phys_qol_model_adh, conf.int = TRUE) %>% filter(term == "hard_drugs_B") %>% mutate(Outcome = "Physical QoL", Group = "With Adherence")) %>%
  mutate(Beta = round(estimate, 2),
         CI_freq = paste0("(", round(conf.low, 2), ", ", round(conf.high, 2), ")"),
         pval = ifelse(p.value < 0.001, "<0.001", as.character(round(p.value, 3)))) %>%
  select(Outcome, Beta, CI_freq, pval)

# Bayesian Results
# Calculating additional posterior probabilities
pp_vload_adh <- calc_post_prob(vload_fit_adh)
pp_tcell_adh <- calc_post_prob(tcell_fit_adh)
pp_ment_adh <- calc_post_prob(ment_qol_fit_adh)
pp_phys_adh <- calc_post_prob(phys_qol_fit_adh)

bayes_adh <- bind_rows(posterior_summary_vload %>% filter(Parameter == "beta[7]") %>% mutate(Outcome = "Viral Load"),
                       posterior_summary_tcell %>% filter(Parameter == "beta[7]") %>% mutate(Outcome = "T-Cell Count"),
                       posterior_summary_ment_qol %>% filter(Parameter == "beta[7]") %>% mutate(Outcome = "Mental QoL"),
                       posterior_summary_phys_qol %>% filter(Parameter == "beta[7]") %>% mutate(Outcome = "Physical QoL"),
                       posterior_summary_vload_adh %>% filter(Parameter == "beta[7]") %>% mutate(Outcome = "Viral Load"),
                       posterior_summary_tcell_adh %>% filter(Parameter == "beta[7]") %>% mutate(Outcome = "T-Cell Count"),
                       posterior_summary_ment_qol_adh %>% filter(Parameter == "beta[7]") %>% mutate(Outcome = "Mental QoL"),
                       posterior_summary_phys_qol_adh %>% filter(Parameter == "beta[7]") %>% mutate(Outcome = "Physical QoL")) %>%
  mutate(Post_Mean = round(Estimate, 2),
         CI_bayes = paste0("(", round(HPDI_2.5, 2), ", ", round(HPDI_97.5, 2), ")"),
         Post_Prob = c(pp_vload, pp_tcell, pp_ment, pp_phys,
                       pp_vload_adh, pp_tcell_adh, pp_ment_adh, pp_phys_adh)) %>%
  select(Outcome, Post_Mean, CI_bayes, Post_Prob)


final_table <- bind_cols(combined_adh, bayes_adh %>% select(-Outcome))

# Create table
final_table %>%
  kbl(col.names = c("Outcome", "β", "95% CI", "p-value", "Posterior Mean", "95% HPDI", "P(β < 0)"),
      align = c("l", "r", "r", "r", "r", "r", "r")) %>%
  kable_classic(full_width = FALSE) %>%
  add_header_above(c(" " = 1, "Frequentist" = 3, "Bayesian" = 3)) %>%  # changed 2 to 3
  pack_rows("Without Adherence", 1, 4) %>%
  pack_rows("With Adherence", 5, 8) %>%
  footnote(general = "Viral load was log10 transformed prior to analysis",
           general_title = "",
           footnote_as_chunk = TRUE)

```






