---
title: "Project 1 - Preliminary Work"
author: "Madelynn Schina"
date: "February 2026"
output: pdf_document
---


```{r, include=F} 
knitr::opts_chunk$set(echo = TRUE)

library(knitr)
library(here)
library(dplyr)
library(ggplot2)
library(gtsummary)

```



```{r, echo = F}
# Loading dataset
hiv_data <- read.csv(here("Project 1/Data", "hiv_6624_final.csv"))

```


Questions of Interest: How treatment response 2 years after initiating HAART differs between subjects who report using hard drugs, such as heroin and cocaine, at baseline and other subjects, who did not report hard drug use at baseline.


Variables of Interest:

Outcomes: VLOAD, LEU3N
Primary Predictors: hard_drugs at baseline, ADH (adherence to treatment)
Adjusting Covariates: age, race, bmi



Questions for the PI:
- When you say you are interested in treatment response, are you only interested in you laboratory measures or quality of life measures as well? - interested in all four outcomes of interest mentioned
- Any limitations on values of viral load or t-cell count? Anything biologically implausible or a limitation of the assay? - 
- Do men smoking and/or drinking react differently to treatment? Should these covariates be accounted for? Do you want to adjust for kidney and liver disease as they impact the immune system? -- Yes, No, No, No
- Am I correct that syphilis, herpes and gonorrhea can all exacerbate HIV symptoms? Saw these indicators in the dictionary but not the dataset, do you not want them adjusted for? -- No


- hard_drugs at baseline, nothing after
- adherence, race, educ (not income tied together)
- consider collapsing race
- could hard_drug explain difference in adherence (overarching question)
- refuse to answer should be considered as missing
- vif to test multicolinarity of years and age



```{r, echo = F}
hiv_data %>%
  select(hard_drugs, ADH, RACE, age, BMI, AGG_MENT, AGG_PHYS) %>%
  tbl_summary(statistic = list(all_continuous() ~ "{mean} ({sd})",
                               all_categorical() ~ "{n} ({p}%)"),
    missing = "ifany")


hiv_data %>%
  filter(years %in% c(0, 1, 2)) %>%
  select(hard_drugs, ADH, RACE, age, BMI, AGG_MENT, AGG_PHYS) %>%
  tbl_summary(statistic = list(all_continuous() ~ "{mean} ({sd})",
                               all_categorical() ~ "{n} ({p}%)"),
    missing = "ifany")


```



```{r, echo = F}
# Histogram of Viral Load
ggplot(hiv_data, aes(x = VLOAD)) +
  geom_histogram(fill = "cornflowerblue", color = "black") +
  labs(x = "Viral Load",
       y = "Count")

# Histogram of T-Cell Count
ggplot(hiv_data, aes(x = LEU3N)) +
  geom_histogram(fill = "cornflowerblue", color = "black", binwidth = 50) +
  labs(x = "T-Cell Count",
       y = "Count")

# Histograms of QOL, Mental and Physical
ggplot(hiv_data, aes(x = AGG_MENT)) +
  geom_histogram(fill = "cornflowerblue", color = "black") +
  labs(x = "QOL Mental",
       y = "Count")

ggplot(hiv_data, aes(x = AGG_PHYS)) +
  geom_histogram(fill = "cornflowerblue", color = "black") +
  labs(x = "QOL Mental",
       y = "QOL Physical")

```



Personal Notes:
- Focused on baseline and 2-year visit
- Remember to mention the type of prior used in your report (listed in slide deck), should also mention assumptions made (also in slide deck)
- Either center biological data or move center (biological data typically doesn't have an intercept of 0, implausible)





