---
title: "Project 0 - Report Analysis"
author: "Madelynn Schina"
date: "January 2026"
output: pdf_document
---


```{r, include=F}
knitr::opts_chunk$set(echo = TRUE)

library(knitr)
library(here)
library(dplyr)
library(tidyr)
library(ggplot2)
library(lmerTest)
library(gtsummary)

```


```{r, echo = F}
# Loading dataset
hormone_data <- read.csv(here("Project 0/Data", "Project0_Clean_v2.csv"))

```



```{r, echo = F}
## Data Cleaning
# Removing any observations with recorded cortisol levels above 80 (biological impossible and thus a lab error)
removed_count_cort <- hormone_data %>%
  filter(Cortisol..nmol.L. >= 80) %>%
  nrow()

hormone_data_clean <- hormone_data %>%
  filter(Cortisol..nmol.L. < 80)

# One observation found with a cortisol measurement over 80
# Five observations missing cortisol measurements, error code present

# Removing any observations with recorded DHEA of 5.205, limitation of instrument
removed_count_dhea <- hormone_data_clean %>%
  filter(DHEA..nmol.L. == 5.205) %>%
  nrow()

hormone_data_clean <- hormone_data_clean %>%
  filter(DHEA..nmol.L. != 5.205)

# Six observations with a DHEA measurement of 5.205


# 12 observations total removed due to measurement criteria

```




# Question 1
Question: What is the agreement between the subject’s recordings of sampling times compared to the times recorded by an electronic monitoring cap?


```{r, echo = F}
## Handling missing data and variable formatting to answer question 1
# Removing any observations missing booklet or cap times
removed_count_time <- hormone_data_clean %>%
  filter(if_any(all_of(c("Booket..Clock.Time", "MEMs..Clock.Time")), ~ . == "")) %>%
  nrow()

data_clean_one <- hormone_data_clean %>%
  filter(if_all(all_of(c("Booket..Clock.Time", "MEMs..Clock.Time")), ~ . != ""))

# 83 observations missing either or both times



# Ensuring all necessary times are treated correctly
# Converting them into minutes for the sake of future calculations
data_clean_one <- data_clean_one %>%
  separate(Booket..Clock.Time, into = c("h1", "m1"), sep = ":", convert = TRUE) %>% # booklet time
  separate(MEMs..Clock.Time, into = c("h2", "m2"), sep = ":", convert = TRUE) %>% # cap times
  mutate(booklet_time_min = h1 * 60 + m1,
         cap_time_min = h2 * 60 + m2)


# Calculating the difference between booklet and cap time
data_clean_one <- data_clean_one %>%
  mutate(diff_min = booklet_time_min - cap_time_min)

```


```{r, echo = F}
# Fitting LMM Model with random intercept for subjects
time_lmm <- lmer(cap_time_min ~ booklet_time_min + (1 | SubjectID), data = data_clean_one)

# Creating summary table of model results for question 1
tbl_regression(time_lmm,
               conf.level = 0.95,
               intercept = TRUE,
               label = list(`(Intercept)` ~ "Intercept",
                            booklet_time_min ~ "Slope"),
               estimate_fun = ~style_number(.x, digits = 2))

```


```{r, echo = F}
# Diagnostic plots of model for question 1
# Residuals vs Fitted
plot(fitted(time_lmm), resid(time_lmm))
abline(h = 0, col = "red")

# Q-Q plot of residuals
qqnorm(resid(time_lmm))
qqline(resid(time_lmm), col = "red")


```




```{r, echo = F, warning = F, message = F}
## Plots for Question 1
# Histogram of differences in times
ggplot(data_clean_one, aes(x = diff_min)) +
  geom_histogram(bins = 30, fill = "cornflowerblue", color = "black") +
  labs(x = "Time Difference (minutes)",
       y = "Count")

# Scatterplot of time points, includes a dashed line with a slope of one and a solid line for the regression model
data_clean_one$fitted <- predict(time_lmm)

ggplot(data_clean_one, aes(x = booklet_time_min, y = cap_time_min)) +
  geom_point(alpha = 0.3, aes(color = "Data Points")) +
  geom_abline(data = data.frame(x = 0, y = 0), aes(intercept = 0, slope = 1, color = "Ideal Fit"),
              linetype = "dashed", linewidth = 1.2) +
  geom_line(aes(y = fitted, color = "LMM Fit"), linewidth = 1.2) +
  scale_color_manual(name = "Legend", values = c("Data Points" = "seagreen",
                                                 "Ideal Fit" = "darkorchid2",
                                                 "LMM Fit" = "cornflowerblue")) +
  labs(x = "Booklet Time (minutes)",
       y = "Cap Time (minutes)",
       title = "Cap Time vs. Booklet Time")

```



Notes/Comments for Write-Up:
- Both plots, the histogram and scatterplot, indicate that the times are pretty close in agreement.     
- The cap time was on average 13 minutes higher than the first, while the relationship between the two measurements was nearly perfectly linear (slope = 0.992), indicating minimal proportional bias but a small systematic offset.





# Question 2
Are subjects adhering accurately to the +30min and +10 hour sampling times required by a study protocol?


```{r, echo = F, message = F, warning = F}
# Adding wake-up time to each observation of a given day for each individual patient
# Converting wake-up time into minutes
adherence_data <- hormone_data_clean %>%
  group_by(SubjectID, Collection.Date) %>%
  mutate(Sleep.Diary.reported.wake.time = Sleep.Diary.reported.wake.time[1]) %>%
  filter(trimws(Sleep.Diary.reported.wake.time) != "") %>%
  separate(Sleep.Diary.reported.wake.time, into = c("wake_h", "wake_m"), sep = ":", convert = TRUE) %>%
  mutate(wake_min = wake_h * 60 + wake_m) %>%
  ungroup()


# Calculating target times from wake up for samples 2 and 4
adherence_data <- adherence_data %>%
  mutate(target_time = case_when(Collection.Sample == 2 ~ wake_min + 30,
                                 Collection.Sample == 4 ~ wake_min + 600))


# Removing any observations missing booklet time and converting it into minutes
adherence_data <- adherence_data %>%
  mutate(Booket..Clock.Time = na_if(Booket..Clock.Time, "")) %>%
  filter(!is.na(Booket..Clock.Time)) %>%
  separate(Booket..Clock.Time, into = c("h3", "m3"), sep = ":", convert = TRUE) %>%
  mutate(booklet_time_min = h3 * 60 + m3)


# Classifying adherence based on time difference
adherence_data <- adherence_data %>%
  mutate(diff_min = abs(booklet_time_min - target_time),
         adherence_label = case_when(diff_min <= 7.5 ~ "Good",
                                     diff_min <= 15 & diff_min > 7.5 ~ "Adequate",
                                     diff_min > 15 ~ "Poor")) %>%
  mutate(adherence_label = factor(adherence_label, levels = c("Good", "Adequate", "Poor")))


# Creating table of adherence results by sampling times, only interested in samples 2 and 4
adherence_table <- adherence_data %>%
    filter(Collection.Sample %in% c(2, 4)) %>%
  select(adherence_label, Collection.Sample) %>%
  tbl_summary(by = Collection.Sample,
              label = list(adherence_label ~ "Adherence Classification"),
              digits = list(all_categorical() ~ 0)) %>%
  modify_header(label ~ "",
                update = c("stat_1" ~ "30-Minute Sample<br>(N={n})",
                           "stat_2" ~ "600-Minute Sample<br>(N={n})"))

adherence_table

```





# Question 3
What are the changes of cortisol and DHEA over time?


```{r, echo = F}
# Diagnostic plots to see if transformations are required of outcomes
# Histogram of cortisol levels
ggplot(adherence_data, aes(x = Cortisol..nmol.L.)) +
  geom_histogram(bins = 30, fill = "cornflowerblue", color = "black") +
  labs(x = "Cortisol Levels (nmol/L)",
       y = "Count")

# Histogram of DHEA
ggplot(adherence_data, aes(x = DHEA..nmol.L.)) +
  geom_histogram(bins = 30, fill = "seagreen", color = "black") +
  labs(x = "DHEA (nmol/L)",
       y = "Count")

# Histograms suggest that both variables should be log transformed
# Log transforming both variables
time_data <- adherence_data %>%
  mutate(log_cort = log(Cortisol..nmol.L.),
         log_dhea = log(DHEA..nmol.L.))

# Histogram of transformed cortisol levels
ggplot(time_data, aes(x = log_cort)) +
  geom_histogram(bins = 30, fill = "cornflowerblue", color = "black") +
  labs(x = "Cortisol Levels (nmol/L)",
       y = "Count")

# Histogram of transformed DHEA
ggplot(time_data, aes(x = log_dhea)) +
  geom_histogram(bins = 30, fill = "seagreen", color = "black") +
  labs(x = "DHEA (nmol/L)",
       y = "Count")

```


```{r, echo = F}
# Getting time from wake-up to make interpretability of models easier
time_data <- time_data %>%
  mutate(time_since_wake = booklet_time_min - wake_min)

# Creating variables for before and after 30 minutes to create knot to reflect patten we expect to see in the data
time_data <- time_data %>%
  mutate(pre_knot = ifelse(time_since_wake < 30, time_since_wake, 30),
         post_knot = ifelse(time_since_wake >= 30, time_since_wake - 30, 0))



# Fitting piecewise lmm for cortisol levels
cort_model <- lmer(log_cort ~ pre_knot + post_knot + (1 | SubjectID),
                          data = time_data)

# Diagnostic plots for cortisol model
plot(fitted(cort_model), resid(cort_model))
abline(h = 0, col = "red")

qqnorm(resid(cort_model))
qqline(resid(cort_model), col = "red")


# Fitting piecewise lmm for DHEA
dhea_model <- lmer(log_dhea ~ pre_knot + post_knot + (1 | SubjectID),
                          data = time_data)

# Diagnostic plots for DHEA model
plot(fitted(dhea_model), resid(dhea_model))
abline(h = 0, col = "red")

qqnorm(resid(dhea_model))
qqline(resid(dhea_model), col = "red")

```


```{r, echo = F}
# Organizing results for question 3 models
# Results of cortisol model
tbl_cort <- tbl_regression(cort_model,
               conf.level = 0.95,
               intercept = TRUE,
               label = list(`(Intercept)` ~ "Intercept",
                            pre_knot ~ "Waking to 30 minutes",
                            post_knot ~ "After 30 minutes"),
               estimate_fun = ~style_number(.x, digits = 4))

# Results of DHEA model
tbl_dhea <- tbl_regression(dhea_model,
               conf.level = 0.95,
               intercept = TRUE,
               label = list(`(Intercept)` ~ "Intercept",
                            pre_knot ~ "Waking to 30 minutes",
                            post_knot ~ "After 30 minutes"),
               estimate_fun = ~style_number(.x, digits = 5))

# Combining model results into nice table
tbl_combined <- tbl_merge(
  tbls = list(tbl_cort, tbl_dhea),
  tab_spanner = c("**Cortisol Model**", "**DHEA Model**")
)

tbl_combined


```


```{r, echo = F}
## Plots for Question 3
# Getting predictions from fitted models
time_data <- time_data %>%
  mutate(fitted_cort = predict(cort_model, re.form = NA),
         fitted_dhea = predict(dhea_model, re.form = NA))

# Scatterplot for cortisol levels
ggplot(time_data, aes(x = time_since_wake)) +
  geom_point(aes(y = Cortisol..nmol.L., col = 'Observed'), alpha = 0.4) +
  geom_line(aes(y = exp(fitted_cort), col = 'Predicted'), linewidth = 1) +
  scale_color_manual(name = "Legend",
                     values = c("Observed" = "black", "Predicted" = "darkolivegreen3")) +
  theme_bw() +
  labs(x = "Time Since Waking (Minutes)",
       y = "Cortisol (nmol/L)",
       title = "Estimated Mean of Cortisol Over Time") +
  theme(legend.position = 'bottom')

# Scatterplot for DHEA levels
ggplot(time_data, aes(x = time_since_wake)) +
  geom_point(aes(y = DHEA..nmol.L., col = 'Observed'), alpha = 0.4) +
  geom_line(aes(y = exp(fitted_dhea), col = 'Predicted'), linewidth = 1) +
  scale_color_manual(name = "Legend",
                     values = c("Observed" = "black", "Predicted" = "darkcyan")) +
  theme_bw() +
  labs(x = "Time Since Waking (Minutes)",
       y = "DHEA (nmol/L)",
       title = "Estimated Mean of DHEA Over Time") +
  theme(legend.position = 'bottom')

```



# Additional Plots & Tables for Report

```{r, echo = F}
# Table1 Summary for results section of final report
summary_data <- hormone_data_clean %>%
  separate(Booket..Clock.Time, into = c("h1", "m1"), sep = ":", convert = TRUE) %>% # booklet time
  separate(MEMs..Clock.Time, into = c("h2", "m2"), sep = ":", convert = TRUE) %>% # cap times
  mutate(booklet_time_min = h1 * 60 + m1,
         cap_time_min = h2 * 60 + m2) %>%
  select(Collection.Sample, Cortisol..nmol.L., DHEA..nmol.L., booklet_time_min, cap_time_min)

# Renaming Collection.Sample for table
summary_data <- summary_data %>%
  mutate(Collection.Sample = factor(Collection.Sample,
                                    levels = c(1, 2, 3, 4),
                                    labels = c("Wake", "+30 min", "Lunch", "+600 min")))

tbl_summary(summary_data,
            by = Collection.Sample,
            missing = "ifany",
            statistic = all_continuous() ~ "{mean} ({sd})",
            label = list(Cortisol..nmol.L. ~ "Cortisol (nmol/L)",
                         DHEA..nmol.L. ~ "DHEA (nmol/L)",
                         booklet_time_min ~ "Booklet Time (min)",
                         cap_time_min ~ "Cap Time (min)")) %>%
  add_overall() %>%
  modify_table_body(~ .x %>%
                      dplyr::mutate(label = ifelse(label == "Unknown", "Missing", label)))


```





