---
title: "Project 0 - Report Analysis"
author: "Madelynn Schina"
date: "January 2026"
output: pdf_document
---


```{r, include=F}
knitr::opts_chunk$set(echo = TRUE)

library(knitr)
library(here)
library(dplyr)
library(tidyr)
library(ggplot2)
library(lmerTest)
library(gtsummary)

```


```{r, echo = F}
# Loading dataset
horomone_data <- read.csv(here("Project 0/Data", "Project0_Clean_v2.csv"))

```



```{r, echo = F}
## Data Cleaning
# Removing any observations with recorded cortisol levels above 80 (biological impossible and thus a lab error)
removed_count_cort <- horomone_data %>%
  filter(Cortisol..nmol.L. >= 80) %>%
  nrow()

horomone_data_clean <- horomone_data %>%
  filter(Cortisol..nmol.L. < 80)

# One observation found with a cortisol measurement over 80
# Five observations missing cortisol measurements, error code present

# Removing any observations with recorded DHEA of 5.205, limitation of instrument
removed_count_dhea <- horomone_data_clean %>%
  filter(DHEA..nmol.L. == 5.205) %>%
  nrow()

horomone_data_clean <- horomone_data_clean %>%
  filter(DHEA..nmol.L. != 5.205)

# Six observations with a DHEA measurement of 5.205


# 12 observations removed due to measurement criteria

```



# Question 1
Question: What is the agreement between the subject’s recordings of sampling times compared to the times recorded by an electronic monitoring cap?


```{r, echo = F}
## Handling missing data and variable formatting to answer question 1
# Removing any observations missing booklet or cap times
removed_count_time <- horomone_data_clean %>%
  filter(if_any(all_of(c("Booket..Clock.Time", "MEMs..Clock.Time")), ~ . == "")) %>%
  nrow()

data_clean_one <- horomone_data_clean %>%
  filter(if_all(all_of(c("Booket..Clock.Time", "MEMs..Clock.Time")), ~ . != ""))

# 83 observations missing either or both times

# Ensuring all necessary times are treated correctly
# Converting them into minutes for the sake of future calculations
data_clean_one <- data_clean_one %>%
  separate(Booket..Clock.Time, into = c("h1", "m1"), sep = ":", convert = TRUE) %>% # booklet time
  separate(MEMs..Clock.Time, into = c("h2", "m2"), sep = ":", convert = TRUE) %>% # cap times
  mutate(booklet_time_min = h1 * 60 + m1,
         cap_time_min = h2 * 60 + m2)


# Calculating the difference between booklet and cap time
data_clean_one <- data_clean_one %>%
  mutate(diff_min = booklet_time_min - cap_time_min)

```


```{r, echo = F}
# Fitting LMM Model with random intercept for subjects
time_lmm <- lmer(cap_time_min ~ booklet_time_min + (1 | SubjectID), data = data_clean_one)
summary(time_lmm)

```


```{r, echo = F, warning = F, message = F}
## Plots for Question 1
# Histogram of differences in times
ggplot(data_clean_one, aes(x = diff_min)) +
  geom_histogram(bins = 30, fill = "cornflowerblue", color = "black") +
  labs(x = "Time Difference (minutes)",
       y = "Count")

# Scatterplot of cap time vs. booklet time, includes line with a slope of one
ggplot(data_clean_one, aes(x = booklet_time_min, y = cap_time_min)) +
  geom_point(alpha = 0.6, color = "darkorchid2") +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
  labs(x = "Booklet Time (minutes)", y = "Cap Time (minutes)")

# Scatterplot of time points, includes a dashed line with a slope of one and a solid line for the regression model
ggplot(data_clean_one, aes(x = booklet_time_min, y = cap_time_min)) +
  geom_point(alpha = 0.3, color = "seagreen") +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "darkorchid2", linewidth = 1.2) +
  geom_smooth(method = "lm", se = FALSE, color = "cornflowerblue", linewidth = 0.8) +
  labs(
    x = "Booklet Time (minutes)",
    y = "Cap Time (minutes)",
    title = "Agreement of Time Measurements"
  )

```



Notes/Comments for Write-Up:
- Both plots, the histogram and scatterplot, indicate that the times are pretty close in agreement.     
- The cap time was on average 13 minutes higher than the first, while the relationship between the two measurements was nearly perfectly linear (slope = 0.992), indicating minimal proportional bias but a small systematic offset.





# Question 2
Are subjects adhering accurately to the +30min and +10 hour sampling times required by a study protocol?


```{r, echo = F, message = F, warning = F}
# Adding wake-up time to each observation of a given day for each individual patient
# Converting wake-up time into minutes
adherence_data <- horomone_data_clean %>%
  group_by(SubjectID, Collection.Date) %>%
  mutate(Sleep.Diary.reported.wake.time = Sleep.Diary.reported.wake.time[1]) %>%
  filter(trimws(Sleep.Diary.reported.wake.time) != "") %>%
  separate(Sleep.Diary.reported.wake.time, into = c("wake_h", "wake_m"), sep = ":", convert = TRUE) %>%
  mutate(wake_min = wake_h * 60 + wake_m) %>%
  ungroup()


# Subsetting data to focus on just samples 2 and 4 & Calculating target times from wake up
adherence_data <- adherence_data %>%
  filter(Collection.Sample %in% c(2, 4)) %>%
  mutate(target_time = case_when(Collection.Sample == 2 ~ wake_min + 30,
                                 Collection.Sample == 4 ~ wake_min + 600))


# Handling missing times to exclude the least amount of observations possible
# Default is booklet time. If it is missing, will use cap time. If both missing, observation will be removed from analysis
adherence_data <- adherence_data %>%
  mutate(Booket..Clock.Time = na_if(Booket..Clock.Time, ""),
         MEMs..Clock.Time   = na_if(MEMs..Clock.Time, "")) %>%
  mutate(sample_time = coalesce(Booket..Clock.Time, MEMs..Clock.Time)) %>%
  filter(!is.na(sample_time))

# Converting sample_time into minutes for the sake of future calculations
adherence_data <- adherence_data %>%
  separate(sample_time, into = c("h3", "m3"), sep = ":", convert = TRUE) %>%
  mutate(sample_time_min = h3 * 60 + m3)


# Classifying adherence based on time difference
adherence_data <- adherence_data %>%
  mutate(diff_min = abs(sample_time_min - target_time),
         adherence_label = case_when(diff_min <= 7.5 ~ "Good",
                                     diff_min <= 15 & diff_min > 7.5 ~ "Adequate",
                                     diff_min > 15 ~ "Poor")) %>%
  mutate(adherence_label = factor(adherence_label, levels = c("Good", "Adequate", "Poor")))

# Creating table of adherence results by sampling times
adherence_table <- adherence_data %>%
  select(adherence_label, Collection.Sample) %>%
  tbl_summary(by = Collection.Sample,
              percent = "row",
              label = list(adherence_label ~ "Adherence Classification")) %>%
  modify_header(label ~ "",
                update = c("stat_1" ~ "+30 min Sample (N={n})",
                           "stat_2" ~ "+10 hour Sample (N={n})"))

```





# Question 3
What are the changes of cortisol and DHEA over time?


```{r, echo = F}
# Diagnostic plots to see if transformations are required
# Histogram of cortisol levels
ggplot(adherence_data, aes(x = Cortisol..nmol.L.)) +
  geom_histogram(bins = 30, fill = "cornflowerblue", color = "black") +
  labs(x = "Cortisol Levels (nmol/L)",
       y = "Count")

# Histogram of DHEA
ggplot(adherence_data, aes(x = DHEA..nmol.L.)) +
  geom_histogram(bins = 30, fill = "seagreen", color = "black") +
  labs(x = "DHEA (nmol/L)",
       y = "Count")

# Histograms suggest that both variables should be log transformed
# Log transforming both variables
adherence_data <- adherence_data %>%
  mutate(log_cort = log(Cortisol..nmol.L.),
         log_dhea = log(DHEA..nmol.L.))

# Histogram of transformed cortisol levels
ggplot(adherence_data, aes(x = log_cort)) +
  geom_histogram(bins = 30, fill = "cornflowerblue", color = "black") +
  labs(x = "Cortisol Levels (nmol/L)",
       y = "Count")

# Histogram of transformed DHEA
ggplot(adherence_data, aes(x = log_dhea)) +
  geom_histogram(bins = 30, fill = "seagreen", color = "black") +
  labs(x = "DHEA (nmol/L)",
       y = "Count")

```


```{r, echo = F}
# Creating knot at 30 minutes
adherence_data <- adherence_data %>%
  mutate(timeknot = ifelse(sample_time_min <= (wake_min + 30), 0, sample_time_min - (wake_min + 30)))

# Fitting piecewise linear model for cortisol levels
cort_model <- lmer(log_cort ~ sample_time_min + timeknot + (1 | SubjectID),
                          data = adherence_data)

# Fitting piecewise linear model for DHEA
dhea_model <- lmer(log_dhea ~ sample_time_min + timeknot + (1 | SubjectID),
                          data = adherence_data)

summary(cort_model)
summary(dhea_model)


```


```{r, echo = F}
## Plots
# Scatterplot for cortisol levels



# Scatterplot for DHEA



```









