---
title: "Project 3 - Data Analysis Plan"
author: "Madelynn Schina"
date: "April 2026"
output: pdf_document
---


```{r, include=F} 
knitr::opts_chunk$set(echo = TRUE)

library(knitr)
library(here)


```


```{r, echo = F}
# Loading dataset
frmgham_data <- read.csv(here("Project 3/Data", "frmgham2.csv"))

```




Variables of Interest: 10-year survival for ages 40, 50, & 60 for each group; average survival with no risk factors and with each and all risk factors (also add one additional interesting profile) (3 age groups, 5 risk scenarios)
Absolutely include: diabetes indicator, systolic blood pressure, age
Potential Covariates to Ttest before inclusion: presence of coronary heart disease, on blood pressure medicine or not, smoking status, total cholesterol, bmi



Notes from Class:
- Separate models for each sex
- Look into correlation between blood pressure and blood pressure meds
- Will need to check the covariates for extreme/unreasonable values during data cleaning
- If stroke variable is 1 at first visit, that means they had a stroke previously (not interested in them, remove them)
- Only interested in data for first 10 years, exclude data that is from after 10 years
- Censor anyone that has an event besides stroke and consider anything that happened outside of the 10 years to be censored
- For the secondary question, we are interested in do diabetes and blood pressure change over time; look only at period no 10 year cutoff
- Could potentially have different covariates in male vs. female models or the same, just need to be able to explain/justify why you do it either way (can directly compare sexes if covariates are the same, cannot directly compare if covariates vary; cannot address differences in effects/collinearity if you keep them the same)
- Conduct variable selection for males and females separately and then decide on covariates
- Complete Case Analysis based on covariates we decide to include, want the most amount of data possible
- If anyone dies in the 10-years, stroke censoring should be time of death
- High blood pressure - >160
- Main goal is comparisons within sexes, not extremely interested in comparisons between/across sexes
- Look at proportional hazards assumptions being met (one sentence in report, plots in appendix)
- To get baseline risks: average continuous variables to get high and low risk, groups should be self-explanatory (not on bp meds is high risk)





```{r, echo = F}
# Removing visits after period 1 and keeping baseline predictors


# Censoring strokes taking place after 10 years


```






