---
title: "Project 3 - Data Analysis Plan"
author: "Madelynn Schina"
date: "April 2026"
output: pdf_document
---


```{r, include=F} 
knitr::opts_chunk$set(echo = TRUE)

library(knitr)
library(here)
library(tidyverse)
library(gtsummary)


```


```{r, echo = F}
# Loading dataset
frmgham_data <- read.csv(here("Project 3/Data", "frmgham2.csv"))

```




Variables of Interest: 10-year survival for ages 40, 50, & 60 for each group; average survival with no risk factors and with each and all risk factors (also add one additional interesting profile) (3 age groups, 5 risk scenarios)
Absolutely include: diabetes indicator, systolic blood pressure, age
Potential Covariates to Ttest before inclusion: presence of coronary heart disease, on blood pressure medicine or not, smoking status, total cholesterol, bmi



Notes from Class:
- Separate models for each sex
- Look into correlation between blood pressure and blood pressure meds
- Will need to check the covariates for extreme/unreasonable values during data cleaning
- If stroke variable is 1 at first visit, that means they had a stroke previously (not interested in them, remove them)
- Only interested in data for first 10 years, exclude data that is from after 10 years
- Censor anyone that has an event besides stroke and consider anything that happened outside of the 10 years to be censored
- For the secondary question, we are interested in do diabetes and blood pressure change over time; look only at period no 10 year cutoff
- Could potentially have different covariates in male vs. female models or the same, just need to be able to explain/justify why you do it either way (can directly compare sexes if covariates are the same, cannot directly compare if covariates vary; cannot address differences in effects/collinearity if you keep them the same)
- Conduct variable selection for males and females separately and then decide on covariates
- Complete Case Analysis based on covariates we decide to include, want the most amount of data possible
- If anyone dies in the 10-years, stroke censoring should be time of death
- High blood pressure - >160
- Main goal is comparisons within sexes, not extremely interested in comparisons between/across sexes
- Look at proportional hazards assumptions being met (one sentence in report, plots in appendix)
- To get baseline risks: average continuous variables to get high and low risk, groups should be self-explanatory (not on bp meds is high risk)





```{r, echo = F}
# Checking stroke variable to ensure everyone has one value like we expect
stroke_consistency_check <- frmgham_data %>%
  group_by(RANDID) %>%
  summarise(n_unique_stroke_values = n_distinct(STROKE)) %>%
  filter(n_unique_stroke_values > 1)

# Removing anyone that reported stroke at baseline as it thus took place before the study
frmgham_data_clean <- frmgham_data %>%
  filter(!(RANDID %in% (
    frmgham_data %>%
      filter(STROKE == 1 & TIMESTRK == 0) %>%
      pull(RANDID)
  )))

# Getting number of removed
n_baseline_stroke <- frmgham_data %>%
  filter(STROKE == 1 & TIMESTRK == 0) %>%
  distinct(RANDID) %>%
  nrow()


# Censoring stroke and death at 10 years
# stroke_10yr: indicator of if stroke occurred before 10 years and before death
# stroke_time: if stroke occurred then stroke time, if death occurred before stroke then death time, otherwise 10 years
ten_years <- 365.25 * 10
frmgham_data_subset <- frmgham_data_clean %>%
  mutate(stroke_10yr = ifelse(STROKE == 1 & TIMESTRK <= ten_years & (is.na(TIMEDTH) | TIMESTRK <= TIMEDTH), 1, 0),
         stroke_time = case_when(STROKE == 1 & TIMESTRK <= ten_years & (is.na(TIMEDTH) | TIMESTRK <= TIMEDTH) ~ TIMESTRK,
                                 DEATH == 1 & TIMEDTH <= ten_years ~ TIMEDTH,
                                 TRUE ~ ten_years))

# Checking that censoring worked as expected
censor_check <- frmgham_data_subset %>%
  summarise(n_strokes_after_10yr = sum(stroke_10yr == 1 & stroke_time > ten_years, na.rm = TRUE),
            max_event_time = max(stroke_time[stroke_10yr == 1], na.rm = TRUE))

# Getting the number of individuals that were censored
n_censored_stroke <- frmgham_data_clean %>%
  filter(STROKE == 1 & TIMESTRK > ten_years) %>%
  distinct(RANDID) %>%
  nrow()

n_censored_death <- frmgham_data_clean %>%
  filter(DEATH == 1 & TIMEDTH > ten_years) %>%
  distinct(RANDID) %>%
  nrow()


# Subsetting censored data to only baseline visits since we are only interested in predictors at baseline
frmgham_data_subset <- frmgham_data_subset %>%
  filter(TIME == 0)

# Checking that there are no duplicates after subsetting to just baseline visits
duplicate_check <- frmgham_data_subset %>%
  summarise(n_rows = n(),n_ids = n_distinct(RANDID))


```



```{r, echo = F}
# Descriptive Statistics of variables of interest in the dataset
summary_data <- frmgham_data_subset %>%
  select(SEX, stroke_10yr, DIABETES, SYSBP, AGE, PREVCHD, BPMEDS, CURSMOKE, TOTCHOL, BMI)

# Renaming ADH, RACE, EDUCBAS, and SMOKE for table
summary_data <- summary_data %>%
  mutate(PREVCHD = factor(PREVCHD,
                          levels = c(0,1),
                          labels = c("Free of Disease",
                                     "Prevalent Disease")),
         BPMEDS = factor(BPMEDS,
                         levels = c(0,1),
                         labels = c("Not Currently Taken",
                                    "Currently Taken")),
         CURSMOKE = factor(CURSMOKE,
                           levels = c(0,1),
                           labels = c("Not Current Smoker",
                                      "Current Smoker")))


male_summary_data   <- summary_data %>% 
  filter(SEX == 1) %>%
  select(-SEX)

female_summary_data <- summary_data %>% 
  filter(SEX == 2) %>%
  select(-SEX)


# Creating summary table
male_descriptive_table <- tbl_summary(male_summary_data,
                                 by = stroke_10yr,
                                 missing = "ifany",
                                 statistic = list(all_continuous() ~ "{mean} ({sd})",
                                                  all_categorical() ~ "{n} ({p}%)"),
                                 label = list(DIABETES ~ "Diabetes",
                                              SYSBP ~ "Systolic Blood Pressure (mmHg)",
                                              AGE ~ "Age (years)",
                                              PREVCHD ~ "Coronary Heart Disease",
                                              BPMEDS ~ "Taking Blood Pressure Medication",
                                              TOTCHOL ~ "Total Cholesterol (mg/dL)",
                                              CURSMOKE ~ "Smoking Status",
                                              BMI ~ "BMI (kg/m^2)")) %>%
  add_overall() %>%
  modify_table_body(~ .x %>%
                      dplyr::mutate(label = ifelse(label == "Unknown", "Missing", label)))

female_descriptive_table <- tbl_summary(male_summary_data,
                                        by = stroke_10yr,
                                        missing = "ifany",
                                        statistic = list(all_continuous() ~ "{mean} ({sd})",
                                                         all_categorical() ~ "{n} ({p}%)"),
                                        label = list(DIABETES ~ "Diabetes",
                                                     SYSBP ~ "Systolic Blood Pressure (mmHg)",
                                                     AGE ~ "Age (years)",
                                                     PREVCHD ~ "Coronary Heart Disease",
                                                     BPMEDS ~ "Taking Blood Pressure Medication",
                                                     TOTCHOL ~ "Total Cholesterol (mg/dL)",
                                                     CURSMOKE ~ "Smoking Status",
                                                    BMI ~ "BMI (kg/m^2)")) %>%
  add_overall() %>%
  modify_table_body(~ .x %>%
                      dplyr::mutate(label = ifelse(label == "Unknown", "Missing", label)))



# Save table as image for report
# gtsave(as_gt(descriptive_table), "descriptivetable.png")



```



