---
title: "Project 2 Final Report - Analysis Work"
author: "Madelynn Schina"
date: "March 2026"
output: pdf_document
---


```{r, include=F} 
knitr::opts_chunk$set(echo = TRUE)

library(knitr)
library(here)
library(ggcorrplot)
library(powertools)

```


Outcomes of Interest:
    Changes in Cognitive Function from baseline to year 1; CVLT Total Correct, CVLT Recognition, Benson Figure, MST, Story Recall
    Changes in Brain Structure from baseline to year 1; Cortical Thickness

Primary Predictor of Interest:
    Peripheral Inflammatory Markers: Cytokine Levels & Chemokine Levels
    
Adjusting Covariates:
    Outcomes at baseline
    Age
    Sex
    APOE Genotype
    

Other Notes:
    - Will be analyzing the aMCI and health control groups together
    - Goal is to have at least 80% power
    - Bonferoni correction is too conservation due to known correlations, also don't recommend FDR correction
    - Don't forget correlation calculations
    - Sample Size = aMCI group and healthy control group = 125 + 50 = 175
    - Write power analysis in past tense
    - Discuss how you go from the data collected by the investigator to the final data set and how that impact of the difference will be assessed
    
    


```{r, echo = F}
# Loading preliminary dataset
prelim_data <- read.csv(here("Project 2/Data", "PrelimData.csv"))

```


```{r, echo = F}
## Correlation Calculations

# Defining predictors and outcomes in preliminary data
cytokine <- prelim_data$IL_6
chemokine <- prelim_data$MCP_1
cvlt_change <- prelim_data$CVLT_CNG3
cort_thickness_change <- prelim_data$CORT_CNG3

# Defining function to calculate various correlations
calculate_correlation <- function(label, x, y){
  test <- cor.test(x, y, use = "complete.obs")
  data.frame(Comparison = label,
             r = round(test$estimate, 3),
             CI_lower = round(test$conf.int[1], 3),
             CI_upper = round(test$conf.int[2], 3),
             p_value = round(test$p.value, 4),
             n = sum(complete.cases(x,y)))
}


# Calculating Correlations
correlations <- rbind(calculate_correlation("Cytokine ~ Chemokine", cytokine, chemokine),                     # Correlation between the two predictors
                      calculate_correlation("CVLT ~ Cortical Thickness", cvlt_change, cort_thickness_change), # Correlation between the two outcomes
                      calculate_correlation("Cytokine ~ CVLT", cytokine, cvlt_change),
                      calculate_correlation("Cytokine ~ Cortical Thickness", cytokine, cort_thickness_change),
                      calculate_correlation("Chemokine ~ CVLT", chemokine, cvlt_change),
                      calculate_correlation("Chemokine ~ Cortical Thickness", chemokine, cort_thickness_change))

kable(correlations, row.names = FALSE)



# Visualizing Correlations
cor_matrix <- cor(prelim_data, use = "pairwise.complete.obs")
rownames(cor_matrix) <- colnames(cor_matrix) <- c("CVLT Test", "Cortical Thickness", "Cytokine", "Chemokine")

ggcorrplot(cor_matrix,
           method = "circle",
           type = "lower",
           lab = TRUE,
           lab_size = 4,
           colors = c("cornflowerblue", "white", "darkolivegreen2"),
           title = "Correlation Matrix",
           ggtheme = theme_minimal())

```
    




```{r, echo = F}
## Power Calculations

## Aim 1
# Correlation Values - range from 0.2 to 0.7 based on preliminary data
r_vals <- seq(0.2, 0.7, by = 0.05)

# Alpha Values - 2 models for 2 aims with 1 predictor of interset each, 6 outcomes, 24 models total
alpha_levels <- c(0.05/4, 0.05/6, 0.05/24)

# Sample size - 175 as defined in grant
N <- 175

# Initializing results
aim1_results <- data.frame()
aim1_mde <- data.frame()

# Calculating powers
for (i in alpha_levels){
  power_vals <- numeric(length(r_vals))
  
  for (j in seq_along(r_vals)){
    power_vals[j] <- corr.1samp(N = N, rhoA = r_vals[j], alpha = i, sides = 2)
  }
  
  aim1_results <- rbind(aim1_results, data.frame(r = r_vals,
                                                 alpha = i,
                                                 power = power_vals))
}

# Finding MDE
for (i in alpha_levels) {
  
  # Calculate power for each r value
  power_search <- sapply(r_vals, function(r) {
    corr.1samp(N = N, rhoA  = r, alpha = i, sides = 2)
  })
  
  # Find the r value closest to 80% power
  mde <- r_vals[which.min(abs(power_search - 0.80))]
  
  aim1_mde <- rbind(aim1_mde, data.frame(alpha = i,
                                         min_rhoA = round(mde, 3)))}



```




