---
title: "Project 2 Final Report - Analysis Work"
author: "Madelynn Schina"
date: "March 2026"
output: pdf_document
---


```{r, include=F} 
knitr::opts_chunk$set(echo = TRUE)

library(knitr)
library(here)

```


Outcomes of Interest:
    Changes in Cognitive Function from baseline to year 1; CVLT Total Correct, CVLT Recognition, Benson Figure, MST, Story Recall
    Changes in Brain Structure from baseline to year 1; Cortical Thickness

Primary Predictor of Interest:
    Peripheral Inflammatory Markers: Cytokine Levels & Chemokine Levels
    
Adjusting Covariates:
    Outcomes at baseline
    Age
    Sex
    APOE Genotype
    

Other Notes:
    - Will be analyzing the aMCI and health control groups together
    - Goal is to have at least 80% power
    - Bonferoni correction is too conservation due to known correlations, also don't recommend FDR correction
    - Don't forget correlation calculations
    - Sample Size = aMCI group and healthy control group = 125 + 50 = 175
    - Write power analysis in past tense
    - Discuss how you go from the data collected by the investigator to the final data set and how that impact of the difference will be assessed
    
    


```{r, echo = F}
# Loading preliminary dataset
prelim_data <- read.csv(here("Project 2/Data", "PrelimData.csv"))

```


```{r, echo = F}
## Correlation Calculations

# Defining predictors and outcomes in preliminary data
cytokine <- prelim_data$IL_6
chemokine <- prelim_data$MCP_1
cvlt_change <- prelim_data$CVLT_CNG3
cort_thickness_change <- prelim_data$CORT_CNG3

# Defining function to calculate various correlations
calculate_correlation <- function(label, x, y){
  test <- cor.test(x, y, use = "complete.obs")
  data.frame(Comparison = label,
             r = round(test$estimate, 3),
             CI_lower = round(test$conf.int[1], 3),
             CI_upper = round(test$conf.int[2], 3),
             p_value = round(test$p.value, 4),
             n = sum(complete.cases(x,y)))
}


# Calculating Correlations
correlations <- rbind(calculate_correlation("Cytokine ~ Chemokine", cytokine, chemokine),                     # Correlation between the two predictors
                      calculate_correlation("CVLT ~ Cortical Thickness", cvlt_change, cort_thickness_change), # Correlation between the two outcomes
                      calculate_correlation("Cytokine ~ CVLT", cytokine, cvlt_change),
                      calculate_correlation("Cytokine ~ Cortical Thickness", cytokine, cort_thickness_change),
                      calculate_correlation("Chemokine ~ CVLT", chemokine, cvlt_change),
                      calculate_correlation("Chemokine ~ Cortical Thickness", chemokine, cort_thickness_change))

kable(correlations, row.names = FALSE)

# Due to the high correlation of the two predictors (> 0.7), they should be modeled separately

```
    






